CrystalPlan: an Experiment Planning Tool for Crystallography

### 2010-09-21 docs/CrystalPlan_paper/paper.tex
Refs #182: Added ewald sphere figure and script to generate it

@@

@@

CrystalPlan supports area detectors with flat faces and arbitrary rectangular
dimensions; this handles the detector types in use at SNS (Anger cameras and
8-pack tube detectors), but the software could easily be extended for more
complicated detector shapes as needed. The detector configuration is loaded from
a text file containing the location, size, orientation and number of pixels of
each detector.

@@

CrystalPlan is designed to be able to use a variety of sample orientation
goniometers. A base Goniometer class is subclassed for specific types of
goniometers, allowing a unified API. This base class acts as a \emph{universal
goniometer},
which uses the conventional goniometer angles of
$\phi$, $\chi$, $\omega$ as defined in \cite{busing67}.
Each class inheriting from Goniometer can convert
the desired phi, chi, omega angles to the internal motor positions it requires. The API allows
for limited goniometers: for example, TOPAZ is currently equipped with a
goniometer where chi is fixed at 45 degrees. CrystalPlan handles this by
limiting the available sample orientation angles accessible to the user.

@@

The reciprocal space to be measured is divided into a cubic 3D matrix, the size
of which is determined by the user, who specifies the minimum d-spacing
($d_{min}$, in $\AA$) that he or she is interested in. The resolution of the 3D
matrix is also specified (in $\AA^{-1}$), which determines the number of points to be
modeled. Machine memory limits the resolution that can be achieved.

@@

For each sample orientation, a rotation matrix combining the sample's mounting
orientation (the U matrix loaded from ISAW) and the goniometer rotation
($\Phi$, $\Xi$, $\Omega$) is computed. The coverage in reciprocal space is
computed in this way: for each pixel in the face of a detector, the direction of
the scattered beam is known. Then, two incident beam vectors are considered,
corresponding to the low and high wavelength limits of the detectors and/or
source. The scattered beam direction and the two possible incident beam vectors
are used to to find two q vectors, which point in the same direction. At the
high wavelength we find qmin, and the low wavelength limit scattering off of qmax.
Since the instrument has a wide bandwidth, we fill in the 3D q-space matrix by
filling all points from qmin to the  qmax, for all pixels in the face of the
detector. This quickly defines a volume of coverage, as shown in Figure
\ref{fig:volume_coverage}.

@@

Crystal lattice parameters (lengths $a$, $b$, $c$ and angles $\alpha$, $\beta$,
$\gamma$) are typed in or loaded from the ISAW UB matrix file. CrystalPlan then
generates a list of HKL peaks with d-spacings larger than the $d_{min}$ specified. For each
sample orientation being simulated and for each peak, a function calculates at
what wavelength lambda and in which direction it scatters. If the wavelength is
within range of the limits of the instrument, the beam direction is projected
onto each detector face to determine where, on each detector surface, it will be
measured.

Each HKL peak is a separate object that holds a list of simulated measurements,
including which detector measured it, at what position on the detector face, at
what wavelength, and for which sample orientation. These data will be used to
graphically display coverage and calculate statistics such as redundancy.

@@

For the single reflections, each hkl is tracked separately, but a primary hkl is
found that points to a list of equivalent hkls. The GUI can then display either
all hkl, ignoring crystal symmetry; or combine them to display only part of the
reflections.

@@

Genetic algorithms have been covered in the literature before, see e.g ref
[INSERT REFERENCE HERE]. In this application, the genes consist of goniometer
angles. For instance, if a goniometer has freedom in 3 angles and the user
requests a 10 different sample orientations, then each individual has 30 genes.
The initial population is created randomly and is made to evolve using mutations
and crossover. For each individual set of sample orientations, the coverage
statistics are computed (taking crystal symmetry into account if desired), and
this is used as the fitness of the given individual in the normal genetic
algorithm process.

@@

CrystalPlan was written in Python, a powerful, open, cross-platform scripting
language. Python was chosen for its ease of cross-platform deployment, and for
its attractive language features. The numpy and scipy libraries (two
open-source, commonly installed Python libraries) were used for many of the
calculations. Some of the critical calculations were also written with inline C
code, which speed up some calculations by up to $400\times$ as compared with
the pure Python version (which can still be used, in case of incompatibilities).

@@

\begin{figure}
\caption{volume coverage}
%\includegraphics{fig1.ps}
\label{fig:volume_coverage}
\end{figure}

\
\begin{figure}
\caption{Screenshots of the main GUI}
%\includegraphics{fig1.ps}
\label{fig:screenshots}
\end{figure}

@@

\begin{figure}
\caption{figureeeeee}
%\includegraphics{fig1.ps}
\label{fig:screenshots}
\end{figure}

>>>>

\usepackage[pdftex]{graphicx}

@@

Crystal plan is designed with these capabilities:

\begin{itemize}
  \item Load an instrument geometry.
  \item Calculate the orientation of a crystal at different goniometer settings.
  \item Calculate the coverage of these orientations in reciprocal space.
  \item Take crystal symmetries into account.
  \item Optimize the experiment plan.
\end{itemize}

@@

CrystalPlan supports area detectors with flat faces and arbitrary rectangular
dimensions; this handles the detector types in use at SNS (Anger cameras and
8-pack tube detectors), but the software could easily be extended for more
complicated detector shapes as needed. The detector configuration is loaded from
a text file containing the location, size, orientation and number of pixels of
each detector. Other settings to be set for an instrument include the
detectable range of neutron/x-ray wavelength.

@@

CrystalPlan is designed to be able to use a variety of sample orientation
goniometers. A base Goniometer class is subclassed for specific types of
goniometers, allowing a unified interface. This base class acts as a
\emph{universal goniometer},
which uses the conventional goniometer angles of
$\phi$, $\chi$, $\omega$ as defined in \cite{busing67}.
Each class inheriting from Goniometer can convert
the desired phi, chi, omega angles to the internal motor positions it requires. The API allows
for limited goniometers: for example, TOPAZ is currently equipped with a
goniometer where chi is fixed at 45 degrees. CrystalPlan handles this by
limiting the available sample orientation angles accessible to the user.

@@

The reciprocal space to be measured is divided into a cubic 3D matrix, the size
of which is determined by the user, who specifies the minimum d-spacing
($d_{min}$, in $\AA$) that he or she is interested in. The resolution of the 3D
matrix is also specified (in $\AA^{-1}$), which determines the number of points to be
modeled. Available machine memory limits the resolution that can be achieved in
practice.

@@

For each sample orientation, a rotation matrix combining the sample's mounting
orientation (the U matrix loaded from ISAW) and the goniometer rotation
($\Phi$, $\Xi$, $\Omega$) is computed. The coverage in reciprocal space is
computed in this way: for each pixel in the face of a detector, the direction of
the scattered beam is known. Then, two incident beam vectors are considered,
corresponding to the $\lambda_{min}$ and $\lambda_{max}$ wavelength limits
of the detectors and/or source. The scattered beam direction and the two possible incident beam vectors
are used to to find two $q$ vectors, which point in the same direction. At the
high wavelength we find $q_{min}$, and the low wavelength limit scattering off
of $q_{max}$. Since the instrument has a wide bandwidth, we fill in the 3D q-space
matrix by filling all points from $q_{min}$ to the $q_{max}$, for all pixels in
the face of the detector. This quickly defines a volume of coverage, as shown
schematically in two dimensions in Figure \ref{fig:ewald_fig}.

@@

Crystal lattice parameters (lengths $a$, $b$, $c$ and angles $\alpha$, $\beta$,
$\gamma$) are typed in or loaded from the ISAW UB matrix file. CrystalPlan then
generates a list of $hkl$ peaks with d-spacings larger than the $d_{min}$
specified. For each sample orientation being simulated and for each peak, a function calculates at
what wavelength $\lambda$ and in which direction it scatters. If $\lambda$ 
is within range of the limits of the instrument, the beam direction is projected
onto each detector face to determine where, on each detector surface, it will be
measured.

Each $hkl$ peak is a separate object that holds a list of predicted
measurements, including which detector measured it, at what position on the
detector face, at what wavelength, and for which sample orientation. These data
will be used to  graphically display coverage and calculate statistics such as
redundancy.

@@

For the single reflections, each $hkl$ is tracked separately, but a primary
$hkl$ is found that points to a list of equivalent $hkl$'s. The GUI can then display
either all $hkl$, ignoring crystal symmetry; or combine them to display only part of
the reflections.

@@

Genetic algorithms have been covered in the literature before, see e.g ref
[INSERT REFERENCE HERE]. In this application, the genes consist of goniometer
angles. For instance, if a goniometer has freedom in 3 angles and the user
requests a 10 different sample orientations, then each individual has 30 genes.
The initial population is created randomly and is made to evolve using mutations
and crossover. Individuals whose goniometer angles turn out to be impossible to
reach are eliminated. For each individual set of sample orientations, the
coverage statistics are computed (taking crystal symmetry into account if desired), and this is used as the fitness of the given individual in the normal genetic
algorithm process.

@@

To this end, CrystalPlan was written in Python, a powerful, open, cross-platform
scripting language. Python was chosen for its ease of cross-platform deployment, and for
its attractive language features. The numpy and scipy libraries (two
open-source, commonly installed Python libraries) were used for many of the
calculations. Some of the critical calculations were also written with inline C
code, which speed up some calculations by up to $400\times$ as compared with
the pure Python version (which can still be used, in case of incompatibilities).

\subsection{Code Layout}

@@

\begin{figure}
\caption{Ewald sphere construction showing reciprocal-space detector coverage
for a wide-bandwidth Laue diffractometer. The smaller blue circle represents the
$1/\lambda_{max}$ limit of measurement, with the larger green circle
representing $1/\lambda_{min}$. The origin of reciprocal space (000) is common to both
spheres, but the sample position in the construction (S1 or S2) depends on
$\lambda$. The angular coverage of the detector is represented by the dashed
wedge, and projects onto each sphere of radius $1/\lambda$ an arc. The red wedge
between the two $\lambda$ is built by projecting straight lines between the
$q_{min}$ and $q_{max}$ limits, giving the volume in reciprocal space that is
measured at a single sample oriention. }
%\includegraphics[bb=0in 0in 2in 2in, width=2in]{ewald_fig.pdf} 
\label{fig:ewald_fig} 
\end{figure}

\begin{figure}
\caption{Screenshots of the main GUI}
%\includegraphics{fig1.ps}
\label{fig:screenshots}
\end{figure}

@@
